Between Subgraph Isomorphism and Maximum Common Subgraph

### 2016-09-13 paper/paper.tex
Easy changes first

% cref style
\crefname{figure}{Figure}{Figures}
\Crefname{figure}{Figure}{Figures}

@@

\begin{abstract}
    When a small pattern graph does not occur inside a larger target graph, we can ask how to find
    ``as much of the pattern as possible'' inside the target graph. In general, this is known as the maximum
    common subgraph problem, which is much more computationally challenging in practice than
    subgraph isomorphism. We introduce a restricted alternative, where we ask if all but $k$
    vertices from the pattern can be found in the target graph. This allows for the development of
    slightly weakened forms of certain invariants from subgraph isomorphism which are based upon degree and
    number of paths.  We show that when $k$ is small, weakening the invariants still retains much of
    their effectiveness. We are then able to solve this problem on the standard SIP problem
    instances used to benchmark subgraph isomorphism algorithms, despite these instances being
    too large for current maximum common subgraph algorithms to handle. Finally, by iteratively
    increasing $k$, we obtain an algorithm which is also competitive for the maximum common subgraph
    problem.
\end{abstract}

@@

The subgraph isomorphism problem is to find a copy of a small \emph{pattern} graph inside a larger
\emph{target} graph. It comes in two forms: in the non-induced variant, edges must be mapped to
edges, but the target may have ``extra edges'', whilst in the induced variant, non-edges may only be
mapped to non-edges. When a pattern cannot be found, we may wish to be given a result which maps as
many vertices of the pattern into the target as possible. In the induced case, this is known as the
maximum common induced subgraph problem (we discuss the non-induced case below). However, although
recent subgraph isomorphism algorithms are comfortable working with graphs with thousands of
vertices, the state of the art for the maximum common subgraph problem
\citep{DBLP:conf/cp/McCreeshNPS16} becomes computationally infeasible at only 35 vertices when
working with unlabelled graphs. This is largely because strong inference, based upon the degrees of
vertices \citep{DBLP:journals/ai/Solnon10} and the distances or paths between them
\citep{DBLP:conf/cp/AudemardLMGP14,DBLP:conf/cp/McCreeshP15}, is possible with subgraph isomorphism,
but not maximum common subgraph, and so the state space in the former is much more restricted,
whilst filtering during search is vastly stronger.

@@

    \end{tikzpicture}
    \caption{The pattern graph $P$ on the left cannot be found in the target graph $T$, but if
    the central vertex in the diamond is removed, then a subgraph isomorphism exists.}
    \label{fig:subgraphexample}
\end{figure}
\subsection{Definitions and Notations}

A non-induced subgraph isomorphism from a graph $P$ to a graph $T$ is an injective mapping $P
\rightarrowtail T $ which maps adjacent vertices to adjacent vertices. An induced subgraph
isomorphism $P \hookrightarrow T$ additionally maps non-adjacent vertices to non-adjacent vertices.
We define a $k$-less-subgraph isomorphism from $P$ to $T$ to be a subgraph isomorphism from all but
$k$ vertices of $P$ to $T$; this may be non-induced or induced, written $P \lessnonind{k} T$ and $P
\lessind{k} T$ respectively.

@@

\begin{proposition}
    \label{prop:comp}
Let $i$ be an assignment of $T$ vertices to $P$ vertices.  Then $i$ satisfies the definition of $P
\hookrightarrow T$ if and only if $i$ satisfies both the definition of $P \rightarrowtail T$ and
of $\loopcomp{P} \rightarrowtail \loopcomp{T}$.
Similarly, $i$ satisfies the definition of $P
\lessind{k} T$ if and only if $i$ satisfies both the definition of $P \lessnonind{k} T$ and of
$\loopcomp{P}
\lessnonind{k} \loopcomp{T}$.
\end{proposition}

A common induced subgraph of graphs $G$ and $H$ is a graph $P$, together with two induced subgraph
isomorphisms to $G$ and $H$; a maximum common induced subgraph is a common induced subgraph with as
many vertices as possible. An induced $k$-less subgraph isomorphism $P \lessind{k} T$ is equivalent
to a common induced subgraph of $P$ and $T$ with $\left|\V(P)\right| - k$ vertices.

@@

        \path (p) edge (a) edge (c)
            (a) edge (d)
            (c) edge (d);
    \end{tikzpicture}
    \caption{There is a non-induced isomorphism $P\rightarrowtail T_{1}$ from the first graph to the
    second, using the highlighted vertices. No such isomorphism exists to the third graph, but there
    is a $1$-less isomorphism $P\lessnonind{1} T_{2}$ omitting vertex $b$.}\label{fig:klessexample}
\end{figure}
\end{example}

@@

Constraints are used to ensure that adjacency pairs of vertices are mapped to adjacenct pairs of
vertices. For the maximum common subgraph case, any pair of assignments with $\bot$ as either value
is also permitted.  To handle the induced case, rather than the more usual approach of directly
having constraints for non-adjacent pairs of vertices, we will make use of \cref{prop:comp} by
seeking an $i$ which simultaneously satisfies $P \rightarrowtail T$ and $\loopcomp{P}
\rightarrowtail \loopcomp{T}$. Finally, an all different except $\bot$ constraint ensures
injectivity.

@@

A non-induced $k$-less-subgraph isomorphism from $P$ to $T$ is equivalent to a subgraph isomorphism
between $P$ and $T$, with $k$ extra universally-adjacent vertices added to $T$. However, using such
an approach is not ideal, because it would introduce symmetries; for induced $k$-less-subgraph
isomorphisms, an approach based around adding vertices to $T$ cannot work at all.
Another algorithmic approach could be to try each way of removing $k$ vertices from the pattern
graph, and solving the subgraph isomorphism problem in turn. This approach might be feasible for $k
= 1$, although it would involve a lot of duplication of search effort, but for larger values of $k$
the number of searches which would have to be made would grow very quickly.
Finally, for the induced case we could try adapting maximum common subgraph algorithms to solve the
decision problem. However, both the maximum common subgraph algorithm and the clique encoding
require $O(\left|\V(P)\right|^2\left|\V(T)\right|^2)$ memory, which is extremely problematic on the
instances we wish to work with.

@@

For datasets, we use the same 5,725 instances used in a recent work on portfolios of algorithms for
subgraph isomorphism \cite{thelionpaper}. This collection includes randomly generated scale-free
graphs \cite{constraints10}, an assortment of real-world graphs of varying sizes \cite{LV02},
randomly generated graph pairs (using bounded degree, regular mesh, and uniform models; all are
satisfiable), segmented images~\cite{pr15,cviu11}, meshes from modelling 3D objects \cite{cviu11},
and graphs close to the satisfiable-unsatisfiable phase transition
\cite{DBLP:conf/ijcai/McCreeshPT16}; all these instances are available in a simple text
format\footnote{\texttt{http://liris.cnrs.fr/csolnon/SIP.html}}.  Note that many of these instances
are \emph{much} larger than were used in the recent comparison of maximum common subgraph algorithms
by \citet{DBLP:conf/cp/McCreeshNPS16}: for unlabelled graphs, they used pairs of graphs with up to
35 vertices in each, whilst our dataset contains graphs with up to 6,671 vertices. For the CP
algorithm, 1,560 of the instances cannot fit in the amount of RAM we have available---we treat these
instances as having timed out. The situation for the clique encoding is even worse, and 3,653 of
these instances do not fit in RAM. In contrast, for our algorithm, every instance fits comfortably
(at least on modern server-class hardware).

\section{Domain Filtering using Degrees}\label{section:degreefiltering}

@@

\begin{restatable}{proposition}{propdeg}\label{prop:deg}
    Let $p$ be a vertex in $P$ and $t$ a vertex in $T$. For both non-induced and induced
    $k$-less-subgraph isomorphisms, if $p \lessmap{k} t$ then
    $\deg(p) - k \le \deg(t)$.
\end{restatable}

A vertex $q$ which is adjacent to $p$ is called its \emph{neighbour}.  Let $\N(p) = \{ q \in \V(P) :
 p \text{ and } q \text{ are adjacent }\}$ be the set of neighbours of $p$.  The \emph{neighbourhood
degree sequence} (NDS) of $p$, $\nds(p)$, is the (non-ascending) sequence of degrees of its
neighbours.

Let $S = ( s_1 , \ldots , s_n )$ and $T = ( t_1 , \ldots , t_m)$ be two sequences.  We say that $S
\preceq T$ if $n \leq m$ and $\forall s_i \in S$ there exists a distinct $t_j \in T$ with $s_i \leq
t_j$.  When considering a $k$-less-subgraph isomorphism we say that $S \lesspreceq{k}
T$ if $n - k \leq m$ and $\forall s_i \in S \setminus S_k$, where $S_k$ is a subsequence of $S$
containing up to $k$ members of $S$, there exists a distinct $t_j \in T$ with $s_i - k \leq t_j$.

@@

\begin{corollary}
Since both $\nds(p)$ and $\nds(t)$ are non-ascending, without loss of generality we can replace
$S \setminus S_k$ from the definition of $S \lesspreceq{k} T$ with $S[k, n]$.
\end{corollary}

@@

In \cref{figure:ids} we show that this is effective in practice: we show the amount of domain reduction
we can achieve at the top of search by using invariants and a fixed $k$, compared to the search
space size for the maximum common induced subgraph problem. (The results on the non-induced version
show a similar but slightly weaker trend.)

Some other invariants do not translate. For example, another rule which can be effective on regular
graphs involves count the number of neighbours of a vertex which are present in a triangle
\cite{mckay2014practical}. Removing a single vertex can alter this count arbitrarily, so we cannot
make use of this fact.

@@

\begin{corollary}
    For a graph $G$, let $G^{n, \ell}$ be the graph with vertex set $\V(G)$. The
    vertices $p$ and $q$ in $G^{n, \ell}$ are adjacent, if there are at least $n$ simple paths of
    length exactly $\ell$ between $p$ and $q$ in $G$. Then any $k$-less-subgraph isomorphism
    $P \lessnonind{k} T$ induces a new $k$-less-subgraph isomorphism
    $P^{n+k, 2}\lessnonind{k} T^{n, 2}$.
\end{corollary}

@@

To allow for fast propagation, rather than calculating paths dynamically like
\citet{DBLP:conf/cp/AudemardLMGP14}, we follow the approach of \citet{DBLP:conf/cp/McCreeshP15} and
construct \emph{supplemental graphs}, and find a mapping which is simultaneously a non-induced subgraph
isomorphism between every supplemental graph pair. We use paths of length 2, looking at counts of 1,
2, and 3 in the target graph (and so counts of $1 + k$ up to $3 + k$ in the pattern).  We then
investigate whether this leads to new constraints being generated.

By an assignment, we mean considering mapping a pattern vertex $p$ to a target vertex $t$ (and not
$\bot$) which does not violate any loop constraints. An assignment pair is two assignments with
distinct $p$ and distinct $t$, which we say is permitted if it does not violate any adjacency
constraint. We define the permitted assignment pair ratio to be the proportion of assignment pairs
which are permitted.  Finally, we scatter plot the permitted assignment pair ratio with and without
supplemental graphs\footnote{Because of the large sizes of the domains, we randomly sample one
million pairs rather than considering every pair. In some cases, we have nearly a thousand domains,
each with nearly ten thousand values---a complete quadratic calculation involving even a trivial
arithmetic operation on this would take many hours.}.

@@

\begin{algorithm*}[tb]
\DontPrintSemicolon
\begin{multicols}{2}
\nl $\FuncSty{klessSubgraphIsomorphism}$ (\\\hspace*{2em}Graph $\mathcal{P}$, Graph $\mathcal{T}$, Int $k$) $\rightarrow$ Bool \;
\nl \Begin{
    \nl \lIf{$\left|\V(\mathcal{P})\right| + k > \left|\V(\mathcal{T})\right|$}{$\KwSty{return}~\KwSty{false}$\label{line:enough}}
    \nl $L \gets \big[ (\mathcal{P},\ \mathcal{T}), (\loopcomp{\mathcal{P}},
        \loopcomp{\mathcal{T}}) \textnormal{~only if we want induced\label{line:comp}},$
        \\\hspace*{1.0em}$(\mathcal{P}^{1 + k,2},\ \mathcal{T}^{1,2}), (\mathcal{P}^{2 + k,2},\
    \mathcal{T}^{2,2}), (\mathcal{P}^{3 + k,2},\ \mathcal{T}^{3,2}) \big]$ \;
    \nl \ForEach{$v \in \V(\mathcal{P})$}{
        \nl $D_v \gets \V(\mathcal{T})$\;
        \nl \ForEach{$(P,\,T) \in L$}{
            \nl $D_v \gets \{ w \in D_v : $ \\
            \hspace*{4em}$v \sim_P v \Rightarrow w \sim_T w~\wedge$ \\
            \hspace*{4em}$\nds_P(v) \lesspreceq{k} \nds_T(w)\}$ \;
        }
        \nl $D_v \gets D_v \cup k~\textnormal{distinct wildcard values}$
    }
    \nl \lIf{\FuncSty{propagate}(D)}{$\KwSty{return}~\FuncSty{search}(L, D, k)$}
    \nl \lElse{$\KwSty{return}~\KwSty{false}$}
}
\bigskip
\nl $\FuncSty{search}$ (GraphPairs $L$, Domains $D$, Int $k$) $\rightarrow$ Bool \;
\nl \Begin{
    \nl \lIf{$D = \emptyset$}{$\KwSty{return}~\KwSty{true}$}
    \nl $D_v \gets \textnormal{the smallest domain in}~D$ \;
    \nl \ForEach {$v' \in D_v~\textnormal{ordered by static degree in}~\mathcal{T}$\label{line:parforeach}}{
        \nl \If{$v'$ \textnormal{is not the first wildcard we have tried\label{line:sym}}}{
            \nl \KwSty{continue} \;
        }
        \nl $D' \gets \FuncSty{clone}(D)$ \;
        \nl $D'_v \gets \{ v' \} $ \;
        \nl \If{\FuncSty{propagate}(D')}{
            \nl \If{$\FuncSty{search}(L, \{ D \in D' : \left|D\right| > 1 \}, k)$}{
                \nl $\KwSty{return}~\KwSty{true}$ \;
            }
        }
    }
}
\bigskip
\nl $\FuncSty{propagate}$ (GraphPairs $L$, Domains $D$) $\rightarrow$ Bool \;
\nl \Begin{
    \nl \While{\KwSty{true}}{
        \nl \If{\textnormal{no effectively-unit domains (treating all wildcard~values as a single
            value) remain\label{line:effect}}}{
            \nl \If{$\KwSty{not}~\FuncSty{allDifferent}(D)$}{
                \nl $\KwSty{return}~\KwSty{false}$ \;
            }
        }
        \nl \If{\textnormal{no effectively-unit domains remain}}{
            \nl $\KwSty{return}~\KwSty{true}$ \;
        }
        \nl $D_v \gets \textnormal{an effectively-unit domain from}~D$ \;
        \nl $v' \gets $ the single value in $D_v$, or an arbitrary wildcard~value if there is more
        than one \;
        \nl \ForEach{$D_w \in D \setminus \{ D_v \}$}{
            \nl $D_w \gets D_w \setminus \{ v' \}$ \label{line:removev} \;
            \nl \ForEach{$(P,\,T) \in L$}{
                \nl \If{$v \sim_P w$}{
                    \nl $D_w \gets D_w \cap \left(\N_T(v') \cup \textnormal{wildcards}\right)$
                }
            }
            \nl \lIf{$D_w = \emptyset$}{$\KwSty{return}~\KwSty{false}$}
        }
    }
}
\bigskip
\nl $\FuncSty{allDifferent}$ (Domains $D$) $\rightarrow$ Bool \;
\nl \Begin{
\nl $(H,\,A,\,n) \gets (\emptyset,\,\emptyset,\,\emptyset,\,0)$ \;
\nl \ForEach{$D_v \in D$ \textnormal{from smallest to largest\label{line:eachdomain}}}{
    \nl $D_v \gets D_v \setminus H$ \label{line:elimhall} \;
    \nl $(A,\,n) \gets (A \cup D_v,\,n + 1)$ \label{line:acc} \;
    \nl \lIf{$D_v = \emptyset~\KwSty{or}~\left|A\right| < n$}{$\KwSty{return}~\KwSty{false}$\label{line:failhall}}
    \nl \lIf{$\left|A\right| = n$}{$(H,\,A,\,n) \gets (H \cup A,\,\emptyset,\,0)$\label{line:hall}}
}
\nl $\KwSty{return}~\KwSty{true}$ \;
}
\end{multicols}
\vspace*{0.4cm}% bug in algorithm2e + multicol: incorrect space left above the vertical line
\caption{An algorithm for the $k$-less subgraph isomorphism problem.}
\label{algorithm:sip}
\end{algorithm*}

We bring these techniques into a full algorithm, listed as \cref{algorithm:sip}. This is derived
from a subgraph isomorphism due to \citet{DBLP:conf/cp/McCreeshP15}, which in a recent evaluation of
algorithm portfolios \citep{thelionpaper} was the single strongest solver. As well as the
modifications to solve $k$-less and induced problems, we have made two simplifications: we do not
use conflict-directed backjumping, and we do not use iterated label filtering to recalculate
neighbourhood degree sequences if vertices are not present in any target domain after
construction---in preliminary experiments, these techniques were not beneficial for $k > 0$. We also
do not strip isolated vertices, as this is not a valid simplification in the induced case.

@@

\begin{figure*}[tb]
    \centering
    \includegraphics*{gen-graph-runtimes.pdf}
    \caption{In the first two plots, the cumulative number of instances solved over time, for the
        induced and non-induced problems, with different values of $k$. We also show the results of
        iteratively increasing $k$ until a solution is found, and in the induced case, the
        performance of two leading maximum common subgraph algorithms. In the third plot, results
        comparing our algorithm with iteratively increasing $k$ to other approaches on maximum
        common induced subgraph instances.}\label{figure:runtimes}
\end{figure*}

>>>>

% cref style
\crefname{figure}{Figure}{Figures}
\Crefname{figure}{Figure}{Figures}
\crefname{section}{Section}{Sections}
\Crefname{section}{Section}{Sections}
\crefname{proposition}{Proposition}{Propositions}
\Crefname{proposition}{Proposition}{Propositions}

@@

\begin{abstract}
    When a small pattern graph does not occur inside a larger target graph, we can ask how to find
    ``as much of the pattern as possible'' inside the target graph. In general, this is known as the maximum
    common subgraph problem, which is much more computationally challenging in practice than
    subgraph isomorphism. We introduce a restricted alternative, where we ask if all but $k$
    vertices from the pattern can be found in the target graph. This allows for the development of
    slightly weakened forms of certain invariants from subgraph isomorphism which are based upon degree and
    number of paths.  We show that when $k$ is small, weakening the invariants still retains much of
    their effectiveness. We are then able to solve this problem on the standard problem
    instances used to benchmark subgraph isomorphism algorithms, despite these instances being
    too large for current maximum common subgraph algorithms to handle. Finally, by iteratively
    increasing $k$, we obtain an algorithm which is also competitive for the maximum common subgraph
    problem.
\end{abstract}

@@

The subgraph isomorphism problem is to find a copy of a small \emph{pattern} graph inside a larger
\emph{target} graph. It comes in two forms: in the non-induced variant, edges must be mapped to
edges, but the target may have ``extra edges'', whilst in the induced variant, non-edges may only be
mapped to non-edges. When a pattern cannot be found, we may wish to be given a result which maps as
many vertices of the pattern into the target as possible. In the induced case, this is known as the
maximum common induced subgraph problem (we discuss the non-induced case in
\cref{section:definitions}). However, although
recent subgraph isomorphism algorithms are comfortable working with graphs with thousands of
vertices, the state of the art for the maximum common subgraph problem
\citep{DBLP:conf/cp/McCreeshNPS16} becomes computationally infeasible at only 35 vertices when
working with unlabelled graphs. This is largely because strong inference, based upon the degrees of
vertices \citep{DBLP:journals/ai/Solnon10} and the distances or paths between them
\citep{DBLP:conf/cp/AudemardLMGP14,DBLP:conf/cp/McCreeshP15}, is possible with subgraph isomorphism,
but not maximum common subgraph, and so the state space in the former is much more restricted,
whilst filtering during search is vastly stronger.

@@

    \end{tikzpicture}
    \caption{The pattern graph $P$ on the left cannot be found in the target graph $T$, but if
    the central vertex in the diamond is removed, then a subgraph isomorphism exists.}
    \label{fig:subgraphexample}
\end{figure}
\subsection{Definitions and Notations} \label{section:definitions}

A non-induced subgraph isomorphism from a graph $P$ to a graph $T$ is an injective mapping $P
\rightarrowtail T $ which maps adjacent vertices to adjacent vertices. An induced subgraph
isomorphism $P \hookrightarrow T$ additionally maps non-adjacent vertices to non-adjacent vertices.
We define a $k$-less subgraph isomorphism from $P$ to $T$ to be a subgraph isomorphism from all but
$k$ vertices of $P$ to $T$; this may be non-induced or induced, written $P \lessnonind{k} T$ and $P
\lessind{k} T$ respectively.

@@

\begin{proposition}
    \label{prop:comp}
Let $i$ be an assignment of $T$ vertices to $P$ vertices.  Then $i$ satisfies the definition of $P
\hookrightarrow T$ if and only if $i$ satisfies $P \rightarrowtail T$ and
$\loopcomp{P} \rightarrowtail \loopcomp{T}$.
Similarly, $i$ satisfies the definition of $P
\lessind{k} T$ if and only if $i$ satisfies both $P \lessnonind{k} T$ and
$\loopcomp{P}
\lessnonind{k} \loopcomp{T}$.
\end{proposition}

A \emph{common induced subgraph} of graphs $G$ and $H$ is a graph $P$, together with two induced subgraph
isomorphisms to $G$ and $H$; a \emph{maximum common induced subgraph} is a common induced subgraph with as
many vertices as possible. An induced $k$-less subgraph isomorphism $P \lessind{k} T$ is equivalent
to a common induced subgraph of $P$ and $T$ with $\left|\V(P)\right| - k$ vertices.

@@

        \path (p) edge (a) edge (c)
            (a) edge (d)
            (c) edge (d);
    \end{tikzpicture}
    \caption{There is a non-induced isomorphism $P\rightarrowtail T$ from the first graph to the
    second, using the highlighted vertices. No such isomorphism exists to the third graph, but there
    is a $1$-less isomorphism $P\lessnonind{1} U$ omitting vertex $b$.}\label{fig:klessexample}
\end{figure}
\end{example}

@@

Constraints are used to ensure that adjacent pairs of vertices are mapped to adjacenct pairs of
vertices. For the maximum common subgraph case, any pair of assignments with $\bot$ as either value
is also permitted.  To handle the induced case, rather than the usual approach of directly
having constraints for non-adjacent pairs of vertices, we will make use of \cref{prop:comp} by
seeking a mapping $i$ which simultaneously satisfies $P \rightarrowtail T$ and $\loopcomp{P}
\rightarrowtail \loopcomp{T}$. Finally, an ``all different except $\bot$ constraint'' ensures
injectivity.

@@

There are three easy ways we might try to approach this problem.
A non-induced $k$-less subgraph isomorphism from $P$ to $T$ is equivalent to a subgraph isomorphism
between $P$ and $T$, with $k$ extra universally-adjacent vertices added to $T$, and so we could try
solving subgraph isomorphism with a modified target graph. However, using such
an approach is not ideal, because it would introduce symmetries; for induced $k$-less subgraph
isomorphisms, an approach based around adding vertices to $T$ cannot work at all.
Another algorithmic approach could be to try each way of removing $k$ vertices from the pattern
graph, and solving each subgraph isomorphism problem in turn. This approach might be feasible for $k
= 1$, although it would involve a lot of duplication of search effort, but for larger values of $k$
the number of searches which would have to be made would grow very quickly.
Finally, for the induced case we could try adapting maximum common subgraph algorithms to solve the
decision problem. However, both the maximum common subgraph algorithm and the clique encoding
require $O(\left|\V(P)\right|^2\left|\V(T)\right|^2)$ memory \citep{DBLP:conf/cp/McCreeshNPS16},
which is extremely problematic on the instances we wish to work with.

@@

For datasets, we use the same 5,725 instances used in a recent work on portfolios of algorithms for
subgraph isomorphism \cite{thelionpaper}. This collection includes randomly generated scale-free
graphs \cite{constraints10}, an assortment of real-world graphs of varying sizes \cite{LV02},
randomly generated graph pairs (using bounded degree, regular mesh, and uniform models; all are
satisfiable), segmented images~\cite{pr15,cviu11}, meshes from modelling 3D objects \cite{cviu11},
and graphs close to the satisfiable-unsatisfiable phase transition
\cite{DBLP:conf/ijcai/McCreeshPT16}. All of these instances are available in a simple text
format\footnote{\texttt{http://liris.cnrs.fr/csolnon/SIP.html}}.  Note that many of these instances
are \emph{much} larger than were used in the recent comparison of maximum common subgraph algorithms
by \citet{DBLP:conf/cp/McCreeshNPS16}: for unlabelled graphs, they used pairs of graphs with up to
35 vertices in each, whilst our dataset contains graphs with up to 6,671 vertices. For the CP
algorithm, 1,560 of the instances cannot fit in the amount of RAM we have available---we treat these
instances as having timed out. The situation for the clique encoding is even worse, and 3,653 of
these instances do not fit in 64GBytes of RAM. In contrast, for our algorithm, every instance fits
comfortably.  \section{Domain Filtering using Degrees}\label{section:degreefiltering}

@@

\begin{restatable}{proposition}{propdeg}\label{prop:deg}
    Let $p$ be a vertex in $P$ and $t$ a vertex in $T$. For both non-induced and induced
    $k$-less subgraph isomorphisms, if $p \lessmap{k} t$ then
    $\deg(p) - k \le \deg(t)$.
\end{restatable}

A vertex $q$ which is adjacent to $p$ is called its \emph{neighbour}.  Let $\N(p) = \{ q \in \V(P) :
 p~\textnormal{and}~q~\textnormal{are adjacent}\}$ be the set of neighbours of $p$.  The \emph{neighbourhood
degree sequence} (NDS) of $p$, $\nds(p)$, is the (non-ascending) sequence of degrees of its
neighbours.

Let $S = ( s_1 , \ldots , s_n )$ and $T = ( t_1 , \ldots , t_m)$ be two sequences.  We say that $S
\preceq T$ if $n \leq m$ and $\forall s_i \in S$ there exists a distinct $t_j \in T$ with $s_i \leq
t_j$.  When considering a $k$-less subgraph isomorphism we say that $S \lesspreceq{k} T$ if $n - k
\leq m$, and if there exists some subsequence $S_k$ of $S$ containing up to $k$ members such that
$\forall s_i \in S \setminus S_k$, there exists a distinct $t_j \in T$ with $s_i - k \leq t_j$.

@@

\begin{corollary}
Since both $\nds(p)$ and $\nds(t)$ are non-ascending, without loss of generality we can replace
$S \setminus S_k$ from the definition of $S \lesspreceq{k} T$ with the subsequence consisting of all
    but the first $k$ members of $S$.
\end{corollary}

@@

\Cref{figure:ids} demonstrates that this is effective in practice: we show the amount of domain
reduction we can achieve at the top of search by using invariants and a fixed $k$, compared to the
search space size for the maximum common induced subgraph problem. (The results on the non-induced
version show a similar but slightly weaker trend.)

Some other invariants do not translate. For example, another rule which can be effective on regular
graphs involves counting the number of neighbours of a vertex which are present in a triangle
\cite{mckay2014practical}. Removing a single vertex can alter this count arbitrarily, so we cannot
make use of this fact.

@@

\begin{corollary}
    For a graph $G$, let $G^{n, \ell}$ be the graph with vertex set $\V(G)$. The
    vertices $p$ and $q$ in $G^{n, \ell}$ are adjacent if there are at least $n$ simple paths of
    length exactly $\ell$ between $p$ and $q$ in $G$. Then any $k$-less subgraph isomorphism
    $P \lessnonind{k} T$ induces a new $k$-less subgraph isomorphism
    $P^{n+k, 2}\lessnonind{k} T^{n, 2}$.
\end{corollary}

@@

To allow for fast propagation, rather than calculating paths dynamically like
\citet{DBLP:conf/cp/AudemardLMGP14}, we follow the approach of \citet{DBLP:conf/cp/McCreeshP15} and
construct \emph{supplemental graphs}, and find a mapping which is simultaneously a non-induced subgraph
isomorphism between every supplemental graph pair. We use paths of length 2, looking at path counts
of 1, 2, and 3 in the target graph (and so counts of $1 + k$ up to $3 + k$ in the pattern).  We then
investigate whether this leads to new constraints being generated.

By an assignment, we mean considering mapping a pattern vertex $p$ to a target vertex $t$ (and not
$\bot$) which does not violate any loop constraints. An assignment pair is two assignments with
distinct $p$ and distinct $t$, which we say is permitted if it does not violate any adjacency
constraint. We define the permitted assignment pair ratio to be the proportion of assignment pairs
which are permitted.  Finally, in \cref{figure:constraints} we scatter plot the permitted assignment
pair ratio with and without supplemental graphs\footnote{Because of the large sizes of the domains,
we randomly sample one million pairs rather than considering every pair. In some cases, we have
nearly a thousand domains, each with nearly ten thousand values---a complete quadratic calculation
involving even a trivial arithmetic operation on this would take many hours.}.

@@

\begin{algorithm*}[tb]
\DontPrintSemicolon
\begin{multicols}{2}
\nl $\FuncSty{klessSubgraphIsomorphism}$ (\\\hspace*{2em}Graph $\mathcal{P}$, Graph $\mathcal{T}$, Int $k$) $\rightarrow$ Bool \;
\nl \Begin{
    \nl \lIf{$\left|\V(\mathcal{P})\right| + k > \left|\V(\mathcal{T})\right|$}{$\KwSty{return}~\KwSty{false}$\label{line:enough}}
    \nl $L \gets \big[ (\mathcal{P},\ \mathcal{T}), (\loopcomp{\mathcal{P}},
        \loopcomp{\mathcal{T}}) \textnormal{~only if we want induced\label{line:comp}},$
        \\\hspace*{1.0em}$(\mathcal{P}^{1 + k,2},\ \mathcal{T}^{1,2}), (\mathcal{P}^{2 + k,2},\
    \mathcal{T}^{2,2}), (\mathcal{P}^{3 + k,2},\ \mathcal{T}^{3,2}) \big]$ \;
    \nl \ForEach{$v \in \V(\mathcal{P})$}{
        \nl $D_v \gets \V(\mathcal{T})$\;
        \nl \ForEach{$(P,\,T) \in L$}{
            \nl $D_v \gets \{ w \in D_v : $ \\
            \hspace*{4em}$v \sim_P v \Rightarrow w \sim_T w~\wedge$ \\
            \hspace*{4em}$\nds_P(v) \lesspreceq{k} \nds_T(w)\}$ \;
        }
        \nl $D_v \gets D_v \cup k~\textnormal{distinct wildcard values}$
    }
    \nl \lIf{\FuncSty{propagate}(D)}{$\KwSty{return}~\FuncSty{search}(L, D, k)$}
    \nl \lElse{$\KwSty{return}~\KwSty{false}$}
}
\bigskip
\nl $\FuncSty{search}$ (GraphPairs $L$, Domains $D$, Int $k$) $\rightarrow$ Bool \;
\nl \Begin{
    \nl \lIf{$D = \emptyset$}{$\KwSty{return}~\KwSty{true}$}
    \nl $D_v \gets \textnormal{the smallest domain in}~D$ \;
    \nl \ForEach {$v' \in D_v~\textnormal{ordered by static degree in}~\mathcal{T}$\label{line:parforeach}}{
        \nl \If{$v'$ \textnormal{is not the first wildcard we have tried\label{line:sym}}}{
            \nl \KwSty{continue} \;
        }
        \nl $D' \gets \FuncSty{clone}(D)$ \;
        \nl $D'_v \gets \{ v' \} $ \;
        \nl \If{\FuncSty{propagate}(D')}{
            \nl \If{$\FuncSty{search}(L, \{ D \in D' : \left|D\right| > 1 \}, k)$}{
                \nl $\KwSty{return}~\KwSty{true}$ \;
            }
        }
    }
}
\bigskip
\nl $\FuncSty{propagate}$ (GraphPairs $L$, Domains $D$) $\rightarrow$ Bool \;
\nl \Begin{
    \nl \While{\KwSty{true}}{
        \nl \If{\textnormal{no effectively-unit domains remain (treating all wildcard~values as a single
            value)\label{line:effect}}}{
            \nl \If{$\KwSty{not}~\FuncSty{allDifferent}(D)$}{
                \nl $\KwSty{return}~\KwSty{false}$ \;
            }
        }
        \nl \If{\textnormal{no effectively-unit domains remain}}{
            \nl $\KwSty{return}~\KwSty{true}$ \;
        }
        \nl $D_v \gets \textnormal{an effectively-unit domain from}~D$ \;
        \nl $v' \gets $ the single value in $D_v$, or an arbitrary wildcard~value if there is more
        than one \;
        \nl \ForEach{$D_w \in D \setminus \{ D_v \}$}{
            \nl $D_w \gets D_w \setminus \{ v' \}$ \label{line:removev} \;
            \nl \ForEach{$(P,\,T) \in L$}{
                \nl \If{$v \sim_P w$}{
                    \nl $D_w \gets D_w \cap \left(\N_T(v') \cup \textnormal{wildcards}\right)$
                }
            }
            \nl \lIf{$D_w = \emptyset$}{$\KwSty{return}~\KwSty{false}$}
        }
    }
}
\bigskip
\nl $\FuncSty{allDifferent}$ (Domains $D$) $\rightarrow$ Bool \;
\nl \Begin{
\nl $(H,\,A,\,n) \gets (\emptyset,\,\emptyset,\,0)$ \;
\nl \ForEach{$D_v \in D$ \textnormal{from smallest to largest\label{line:eachdomain}}}{
    \nl $D_v \gets D_v \setminus H$ \label{line:elimhall} \;
    \nl $(A,\,n) \gets (A \cup D_v,\,n + 1)$ \label{line:acc} \;
    \nl \lIf{$D_v = \emptyset~\KwSty{or}~\left|A\right| < n$}{$\KwSty{return}~\KwSty{false}$\label{line:failhall}}
    \nl \lIf{$\left|A\right| = n$}{$(H,\,A,\,n) \gets (H \cup A,\,\emptyset,\,0)$\label{line:hall}}
}
\nl $\KwSty{return}~\KwSty{true}$ \;
}
\end{multicols}
\vspace*{0.4cm}% bug in algorithm2e + multicol: incorrect space left above the vertical line
\caption{An algorithm for the $k$-less subgraph isomorphism problem.}
\label{algorithm:sip}
\end{algorithm*}

\Cref{algorithm:sip} integrates these techniques into a full algorithm. This is derived from an
algorithm due to \citet{DBLP:conf/cp/McCreeshP15}, which in a recent evaluation of using portfolios
of subgraph isomorphism algorithms was the single strongest solver portfolios \citep{thelionpaper}.
As well as the modifications to solve $k$-less and induced problems, we have made two
simplifications: we do not use conflict-directed backjumping, and we do not use iterated label
filtering to recalculate neighbourhood degree sequences if vertices are not present in any target
domain after construction---in preliminary experiments, these techniques were not beneficial for $k
> 0$. We also do not strip isolated vertices, as this is not a valid simplification in the induced
case.

@@

\begin{figure*}[tb]
    \centering
    \includegraphics*{gen-graph-runtimes.pdf}
    \caption{In the first two plots we show the cumulative number of instances solved over time, for the
        induced and non-induced problems, with different values of $k$. We also show the results of
        iteratively increasing $k$ until a solution is found, and in the induced case, the
        performance of two leading maximum common subgraph algorithms. In the third plot we show results
        comparing iteratively increasing $k$ with our algorithm to other approaches on maximum
        common induced subgraph instances.}\label{figure:runtimes}
\end{figure*}

\begin{figure*}[tb]
    \centering
    \includegraphics*{gen-graph-which-k-by-family.pdf}
    \caption{The proportion of instances, in different families, which become satisfiable for
    increasing values of $k$. The ``larger'' instances are those where we can prove unsatisfiability for
    $k = 5$, whilst the gap between the top of the bar and the top of the graph is the fraction of
    instances where a timeout was reached for at least one value of $k$.}\label{figure:which-k}
\end{figure*}
